Typing the Outputs

### 2015-08-20 sematics/boundary/boundary.tex
various additions

\newcommand {\colfun} [3] {{#1}:>{#2}~\myatsign~{#3}}

@@

@@

This section defines a few input language fragments for later reference.
Those fragments include ``core'' features (\emph{i.e.} simple ones that present no difficulty for any typing system), 
if expressions, first-order recursive functions, higher-order functions, and finally sum/recursive types.
The two function fragments will be used exclusively.

\begin{framed}
\noindent\textbf{Core Features (Base Types, Products, Circles, Let):}
\begin{mathpar}
\infertypeswor [\rmunit] 	{\ty {\tup{}}\rmunit}								{\cdot} 																	\and
\infertypeswor [int]		{\ty {i} \rmint}									{\cdot}																		\and
\infertypeswor [hold]		{\typestwo {\pause e} \rmint}						{\typesone e \rmint}				 										\and
\infertypeswor [int]		{\ty {b} \rmbool}									{\cdot}																		\and
\infertypeswor [hold]		{\typestwo {\pause e} \rmbool}						{\typesone e \rmbool}				 										\and
\infertypeswor [\times I]	{\ty {\tup{e_1,e_2}}{A\times B}}					{\ty {e_1} A & \ty {e_2} B} 												\and
\infertypeswor [\times E_1]	{\ty {\pio e} A}									{\ty e {A\times B}} 														\and
\infertypeswor [\times E_2]	{\ty {\pit e} B}									{\ty e {A\times B}} 														\and
\infertypeswor [\fut I]		{\typesone {\next e}{\fut A}}						{\typestwo e A} 															\and
\infertypeswor [\fut E]		{\typestwo {\prev e} A}								{\typesone e {\fut A}} 														\and
\infertypeswor [\fut E]		{\ty {\letin x {e_1} {e_2}} B}						{\ty {e_1} A & \ty [,\col x A] {e_2} B } 									\and
\end{mathpar}
\end{framed}

\begin{framed}
\noindent\textbf{If Expressions:}
\begin{mathpar}
\infertypeswor [+ E]		{\ty {\ifthen{e_1}{e_2}{e_3}} A}					{\ty {e_1}\rmbool & \ty{e_2} A & \ty{e_3} A}
\end{mathpar}
\end{framed}

@@

\begin{framed}
\noindent\textbf{Sums and Recursion:}
\begin{mathpar}
\infertypeswor [+ I_1]		{\ty {\inl e} {A + B}}								{\ty e A} 																	\and
\infertypeswor [+ I_2]		{\ty {\inr e} {A + B}}								{\ty e B} 																	\and
\infertypeswor [+ E]		{\ty {\caseof{e_1}{x_2.e_2}{x_3.e_3}} C}			{\ty {e_1}{A+B} & \ty[,\col {x_2} A]{e_2} C & \ty[,\col {x_3} B]{e_3} C}	\and
\infertypeswor [\mu I]		{\ty {\roll e} {\mu \alpha.\tau}}					{\ty e {[\mualphatau / \alpha]\tau}} 										\and
\infertypeswor [\mu E]		{\ty {\unroll e} {[\mualphatau / \alpha]\tau}}		{\ty e \mualphatau} 														\and
\end{mathpar}
\end{framed}

\section{Values and Computations}

It turns out that, for the purposes of splitting, it will be handy to differentiate terms into two classes: 
\emph{values} which are fully reduced, and \emph{computations} which may have some reduction remaining.
Importantly, we want to \emph{explicitly note} when subterms of a computation are actually values,
and so we introduce a new piece of syntax to mark just that inclusion: the underline constructor, written $\exv v$
where $v$ is a term (more precisely, a value).  
Note that the type of the underline is just the type of it's contents, which avoids interaction with the type system.

Because staging and the value/computation distinction are both related to reduction, they will interact non-trivially.
For this reason, we'll first devise a  system for a monostage language, before then considering a staged language.

\subsection{Monostage System}

Priori work has observed the difference between values and computations
via a syntactic distinction (i.e. saying there are two kinds of terms, value-terms and computation-terms).
We instead observe the difference judgementally: there's only one kind of term, 
but there are two judgements ($\valsym$ and $\compsym$) which classify terms as one or the other.

For instance, the aforementioned underline constructor always yields a computation,
but its contents are a value:

\begin{framed}
\noindent\textbf{Underlines:}
\begin{mathpar}
\infer {\exv v \comp} {v \val} \and
\infer {\exv v : A} {v : A} 
\end{mathpar}
\end{framed}

Inversely, intro forms are also computations whenever all of the subterms are.
The elim forms, lets, and ifs are always computations.
Note that because we're studying a call-by-value language, all variables are introduced as values.

\begin{framed}
\noindent\textbf{Standard Computations:}
\begin{mathpar}
\infer {\Gamma \vdash \tup{e_1,e_2} \comp}				{\Gamma \vdash e_1 \comp & \Gamma \vdash e_2 \comp} 	\and
\infer {\Gamma \vdash \inl e \comp}						{\Gamma \vdash e \comp} 								\and
\infer {\Gamma \vdash \inr e \comp}						{\Gamma \vdash e \comp} 								\and
\infer {\Gamma \vdash \roll e \comp}					{\Gamma \vdash e \comp} 										
\end{mathpar}
\end{framed}

>>>>

\newcommand {\colfunsym} {:>}
\newcommand {\colfun} [3] {{#1}\colfunsym{#2}~\myatsign~{#3}}

@@

\section{Various Monostage Formalisms}

In this section, we'll cover a few background formalisms in the context of a monostage language.
They will be extended to a staged language in the next section.
This section is primarily intended to be a reference.

\subsection{First-Order Recursive Functions}

We implement first order recursive functions by introducing a new typing judgement \mbox{$f \colfunsym A \to B$},
which says that $f$ is a function from $A$ to $B$.  
We use a different typing judgement here in order to keep functions and other values distinct.
That is, the $\to$ connective only ever appears to the right of ``$:>$'', and never to the right of a ``$:$".
We also allow let statements to work for functions.
Note how the only way to derive a function type is to directly declare a function, or to assume it as a variable.
\begin{framed}
\noindent\textbf{First-Order Recursive Functions:}
\begin{mathpar}
\infertypeswor [\to I]		{\Gamma \vdash {\fix fx{e_1}} \colfunsym {A \to B}}	{\types [\Gamma, f \colfunsym {A \to B},\col x A] {e_1} B} 							\and
\infertypeswor [\to E]		{\types {\app {e_1}{e_2}} {B}}							{\Gamma \vdash {e_1} \colfunsym {A \to B} & \ty {e_2} A} 					\and
\infertypeswor [\fut E]		{\types {\letin f {e_1} {e_2}} B}				{\Gamma \vdash {e_1} \colfunsym {A \to B} 
																			&\types [\Gamma, f \colfunsym {A \to B}] {e_2} B } 					\and
\end{mathpar}
\end{framed}

\subsection{Values and Computations}

It turns out that, for the purposes of splitting, it will be handy to differentiate terms into two classes: 
\emph{values} which are fully reduced, and \emph{computations} which may have some reduction remaining.
Importantly, we want to \emph{explicitly note} when subterms of a computation are actually values,
and so we introduce a new piece of syntax to mark just that inclusion: the underline constructor, written $\exv v$
where $v$ is a term (more precisely, a value).  
The type of the underline is just the type of it's contents, which avoids interaction with the type system.

Because staging and the value/computation distinction are both related to reduction, they will interact non-trivially.
For this reason, we'll first devise a value system for a monostage language.

Prior work has observed the difference between values and computations
via a syntactic distinction (i.e. saying there are two kinds of terms, value-terms and computation-terms).
We instead observe the difference judgementally: there's only one kind of term, 
but there are two judgements ($\valsym$ and $\compsym$) which classify terms as one or the other.

For instance, the aforementioned underline constructor always yields a computation,
but its contents are a value.

\begin{framed}
\noindent\textbf{Underlines:}
\begin{mathpar}
\infer {\Gamma \vdash \exv v \comp} {\Gamma \vdash v \val} \and
\infer {\Gamma \vdash \exv v : A} {\Gamma \vdash v : A} 
\end{mathpar}
\end{framed}

Inversely, intro forms are also computations whenever all of the subterms are.
The elim forms, {\tt let}s, and {\tt if}s are always computations.
Note that because we're studying a call-by-value language, all variables are introduced as values.

\begin{framed}
\noindent\textbf{Standard Computations:}
\begin{mathpar}
\infer {\Gamma \vdash \tup{e_1,e_2} \comp}				{\Gamma \vdash e_1 \comp & \Gamma \vdash e_2 \comp} 	\and
\infer {\Gamma \vdash \inl e \comp}						{\Gamma \vdash e \comp} 								\and
\infer {\Gamma \vdash \inr e \comp}						{\Gamma \vdash e \comp} 								\and
\infer {\Gamma \vdash \roll e \comp}					{\Gamma \vdash e \comp} 								\and
\infer {\Gamma \vdash \pio e \comp}						{\Gamma \vdash e \comp} 								\and
\infer {\Gamma \vdash \pit e \comp}						{\Gamma \vdash e \comp} 								\and
\infer {\Gamma \vdash \unroll e \comp}					{\Gamma \vdash e \comp} 								\and
\infer {\Gamma \vdash \ifthen{e_1}{e_2}{e_3} \comp}		{\Gamma \vdash e_1 \comp 
														&\Gamma \vdash e_2 \comp 
														&\Gamma \vdash e_3 \comp}	\and
\infer {\Gamma \vdash \letin x {e_1} {e_2} \comp}		{\Gamma \vdash e_1 \comp 
														& \Gamma, x \val \vdash e_2 \comp } 				
\end{mathpar}
\end{framed}

Higher-order functions are the only constructs where value terms have computation subterms.

\begin{framed}
\noindent\textbf{Higher-Order Functions:}
\begin{mathpar}			
\infer {\Gamma \vdash \lam x e \val}		{\Gamma, x \val \vdash e \comp} \and
\infer {\Gamma \vdash \app{e_1}{e_2} \comp}	{\Gamma \vdash e_1 \comp & \Gamma \vdash e_2 \comp}
\end{mathpar}
\end{framed}

A term with a first-order function type, $e \colfunsym A \to B$, can only be a lambda itself or a variable declaration.
Ergo, they are always values and never computations.
For this reason, first-order functions don't really participate in the value/computation system.

\begin{framed}
\noindent\textbf{First-Order Recursive Functions:}
\begin{mathpar}
\infer {\Gamma \vdash \app{e_1}{e_2} \comp}	{\Gamma \vdash e_2 \comp}
\end{mathpar}
\end{framed}

@@

This section defines a few input language fragments for later reference.
Those fragments include ``core'' features (base types, products, $\fut$ types, and lets, which present no difficulty for output typing), 
if expressions, first-order recursive functions, higher-order functions, and finally sum/recursive types.
The two function fragments will be used exclusively.

\begin{framed}
\noindent\textbf{Core Feature Typing:}
\begin{mathpar}
\infer {\Gamma \vdash \exv e : A} {\Gamma \vdash e : A} 																									\and
\infertypeswor [\rmunit] 	{\ty {\tup{}}\rmunit}								{\cdot} 																	\and
\infertypeswor [int]		{\ty {i} \rmint}									{\cdot}																		\and
\infertypeswor [hold]		{\typestwo {\pause e} \rmint}						{\typesone e \rmint}				 										\and
\infertypeswor [int]		{\ty {b} \rmbool}									{\cdot}																		\and
\infertypeswor [hold]		{\typestwo {\pause e} \rmbool}						{\typesone e \rmbool}				 										\and
\infertypeswor [\times I]	{\ty {\tup{e_1,e_2}}{A\times B}}					{\ty {e_1} A & \ty {e_2} B} 												\and
\infertypeswor [\times E_1]	{\ty {\pio e} A}									{\ty e {A\times B}} 														\and
\infertypeswor [\times E_2]	{\ty {\pit e} B}									{\ty e {A\times B}} 														\and
\infertypeswor [\fut I]		{\typesone {\next e}{\fut A}}						{\typestwo e A} 															\and
\infertypeswor [\fut E]		{\typestwo {\prev e} A}								{\typesone e {\fut A}} 														\and
\infertypeswor [\fut E]		{\ty {\letin x {e_1} {e_2}} B}						{\ty {e_1} A & \ty [,\col x A] {e_2} B } 									\and
\end{mathpar}
\end{framed}

\begin{framed}
\noindent\textbf{Core Feature Values/Computations:}
\begin{mathpar}
\infer {\Gamma \vdash \exv e \compo} 					{\Gamma \vdash e \valo} 								\and
\infer {\Gamma \vdash \tup{} \valo}						{\cdot} 												\and
\infer {\Gamma \vdash i \valo}							{\cdot}													\and
\infer {\Gamma \vdash b \valo}							{\cdot}													\and
\infer {\Gamma \vdash \tup{e_1,e_2} \valo}				{\Gamma \vdash e_1 \valo & \Gamma \vdash e_2 \valo} 	\and
\infer {\Gamma \vdash \tup{e_1,e_2} \compo}				{\Gamma \vdash e_1 \compo & \Gamma \vdash e_2 \compo} 	\and
\infer {\Gamma \vdash \pio e \compo}					{\Gamma \vdash e \compo} 								\and
\infer {\Gamma \vdash \pit e \compo}					{\Gamma \vdash e \compo} 								\and
\end{mathpar}
\end{framed}

\begin{framed}
\noindent\textbf{If Expressions:}
\begin{mathpar}
\infertypeswor [+ E]		{\ty {\ifthen{e_1}{e_2}{e_3}} A}	{\ty {e_1}\rmbool & \ty{e_2} A & \ty{e_3} A} \and
\infer {\Gamma \vdash \ifthen{e_1}{e_2}{e_3} \compo}			{\Gamma \vdash e_1 \compo 
																&\Gamma \vdash e_2 \compo 
																&\Gamma \vdash e_3 \compo}
\end{mathpar}
\end{framed}

@@

\begin{framed}
\noindent\textbf{Sums and Recursion:}
\begin{mathpar}
\infertypeswor [+ I_1]		{\ty {\inl e} {A + B}}								{\ty e A} 																	\and
\infertypeswor [+ I_2]		{\ty {\inr e} {A + B}}								{\ty e B} 																	\and
\infertypeswor [+ E]		{\ty {\caseof{e_1}{x_2.e_2}{x_3.e_3}} C}			{\ty {e_1}{A+B} & \ty[,\col {x_2} A]{e_2} C & \ty[,\col {x_3} B]{e_3} C}	\and
\infertypeswor [\mu I]		{\ty {\roll e} {\mu \alpha.\tau}}					{\ty e {[\mualphatau / \alpha]\tau}} 										\and
\infertypeswor [\mu E]		{\ty {\unroll e} {[\mualphatau / \alpha]\tau}}		{\ty e \mualphatau} 														\and
\infer {\Gamma \vdash \inl e \valo}						{\Gamma \vdash e \valo} 								\and
\infer {\Gamma \vdash \inl e \compo}					{\Gamma \vdash e \compo} 								\and
\infer {\Gamma \vdash \inr e \valo}						{\Gamma \vdash e \valo} 								\and
\infer {\Gamma \vdash \inr e \compo}					{\Gamma \vdash e \compo} 								\and
\infer {\Gamma \vdash \roll e \valo}					{\Gamma \vdash e \valo} 								\and
\infer {\Gamma \vdash \roll e \compo}					{\Gamma \vdash e \compo} 								\and
\infer {\Gamma \vdash \unroll e \compo}					{\Gamma \vdash e \compo} 								\and
\infer {\Gamma \vdash \letin x {e_1} {e_2} \compo}		{\Gamma \vdash e_1 \compo 
														& \Gamma, x \valo \vdash e_2 \compo } 	
\end{mathpar}
\end{framed}
